\bf Visualizations for Diagnosing Network Models

### 2019-07-10 paper-child.Rnw
just about done but i think it needs some more oomph in the discussion and a bit more connective tissue between sections.

The lack of a native coordinate system has given rise two primary net-vis methods: the node-link diagram and the adjacency matrix visualization \citep{knuth2000, adjmatpro}. We demonstrate these with a simpe example. Consider the network with five nodes, $\{1,2,3,4,5\}$, connected by five directed edges:  $\{2 \to 4, 3 \to 4, 1 \to 5, 3 \to 5, 5 \to 4\}$ shown in Figure~\ref{fig:toyplots}.

The first net-vis method, the node-link diagram, represents nodes with points in two dimensions and then represents edges by connecting the points with lines. These lines can also have arrows on them indicating the direction of the edge for directed networks, as shown in Figure~\ref{fig:toyplots}. Because there is often no natural placement of the nodes, they are placed in 2D at random, then adjusted with a layout algorithm. For a complete overview of available node-link layout algorithms, we direct the reader to \citet{drawingalgs}. Some commonly used algorithms, such as the Kamada-Kawai (KK) layout \citep{kamadakawai} and the Fruchterman-Reingold (FR) layout \citep{fruchterman-reingold}, are designed to mimic physical systems, drawing the graphs based on the ``forces" connecting them. %Other algorithms use multi-dimensional scaling or properties of the adjacency matrix (e.g. its eigenstructure), to place the nodes \citep{drawingalgs}. 
In Figure~\ref{fig:toyplots}, we use the KK algorithm, and unless otherwise stated, all other node-link diagrams in this paper are laid out using the KK algorithm \citep{kamadakawai}.

@@

\subsubsection{Fitting Models to Data}

@@

\noindent \textbf{Models:} To our example data, we fit three different models. Each model uses a simple rate function, $\alpha_m$, and an objective function with two or three parameters. Though our models are very simple, the model-vis techniques we present can easily be applied to much more complex models. The first model of interest, M1, contains the absolute minimum number of parameters in the objective function $f_i(x)$: 
\begin{equation}\label{eq:m1}
M1: \quad f_i(x) = \beta_1s_{i1}(x) + \beta_2s_{i2}(x),
\end{equation}
where $s_{i1}(x)$ is the density network statistic and $s_{i2}(x)$ is the reciprocity network statistic for actor $i$ at the current network state $x$ (see Figure~\ref{fig:structures}). The second and third models, M2 and M3, contain one additional parameter each in the objective function. We chose the additional parameters because they were significant ($p$-value $<$ 0.05) according to a Wald-type test (\texttt{Wald.RSiena()}) provided in the \texttt{RSiena} software \citep{RSienaTest}. The model M2 contains an actor-level covariate parameter, and the model M3 contains an additional strutural effect in the objective function.
\begin{align}\label{eq:m2m3}
M2: \quad f_i(x,z) & = \beta_1s_{i1}(x) + \beta_2s_{i2}(x) + \beta_3s_{i3}(x, z) \\
M3: \quad f_i(x) & = \beta_1s_{i1}(x) + \beta_2s_{i2}(x) + \beta_4s_{i4}(x),
\end{align}

@@

\noindent \textbf{Varying model settings:} The second method for obtaining collections of models is to vary model settings. We have done this already by choosing three different models M1, M2, M3 to fit to our example data sets. CTMC models, however, are much more flexible than the three simple models we have chosen. Other models could have included different rate functions for each node, where for example the rate of change could depend on drinking behavior if we have reason to believe that girls who drink more change friends more frequently. We could also model the behavior changes, not just the network changes, if we believe that friends influence behavior, via peer pressure.  And as before, this method of gathering a collection of models can be used for any network model, such as the hierarchichal Bayesian network estimation model in the \texttt{R} package \texttt{sna}, which allows the user to vary prior and model parameters \citep{sna}.
% took out this next bit just because it doesn't really add much and we can just put the appendix figure into the previous section. % add more later.

@@

%\subsection{Fitting the same model to the same data}\label{sec:samemodsamedat}
\begin{figure}
\begin{subfigure}[t]{.4\textwidth}
<<distests, fig.height=7.5>>=
levels(simu2$parameter) <- c("alpha[1]","alpha[2]","beta[1]","beta[2]","beta[3]","beta[4]")
filter(simu2, parameter %in% c("beta[3]","beta[4]")) %>% 
ggplot() +
  geom_density(aes(x = estimate)) +
  facet_wrap(~parameter, scales = 'free', labeller= label_parsed, nrow = 2) +
  ThemeNoNet +
  scale_fill_brewer(palette = "Greys") + 
  xlab("estimate value") + 
  theme(legend.position = 'bottom')
@
\caption{\label{fig:beta34} Distribution of fitted $\beta_3, \beta_4$ values.}
\end{subfigure}
\begin{subfigure}[t]{.6\textwidth}
<<img, out.width = "\\linewidth">>=
# correlation for other params. 
simsM2 <- filter(simu2, Model == "M2") %>% spread(parameter, estimate)
corm2 <- cor(simsM2[, 3:7])
simsM3 <- filter(simu2, Model == "M3") %>% spread(parameter, estimate)
corM3 <- cor(simsM3[, 3:7])

knitr::include_graphics("img/gray_corplot-1.png")
@
\caption{\label{fig:corplots} Relationships between parameters common to all three models. \nocite{ggally}}
\end{subfigure}
\caption{\label{fig:distests} Distribution of fitted parameter values for our three models. The inclusion of $\beta_3$ or $\beta_4$ clearly has an effect on the distributions of the rate parameters, $\alpha_1$ and $\alpha_2$, and on the other objective function parameters $\beta_1, \beta_2$.}
\end{figure}

\noindent \textbf{Fitting the model repeatedly:} The simulations required to fit the CTMC network models lead to inherent variability in the fitted values. To examine this variability, we fit models M1, M2, and M3 to the friendship network 1,000 times each. The resulting means and standard deviations of the estimates are given in Table~\ref{tab:meanests}, and the distributions of all parameter estimates are shown in Figure~\ref{fig:distests}. In Figure~\ref{fig:corplots}, we can see that the addition of $\beta_3$ to the model results in very different estimates the of the other four parameters, $\alpha_1$, $\alpha_2$, $\beta_1$, and $\beta_2$. In fact, the distributions of $\alpha_1$ and $\beta_1$ for M2 do not overlap at all with the distribution of the same parameters from model M1. On the other hand, the inclusion of the indirect tie parameter, $\beta_4$, does not lead to such drastically different estimates of other model parameters. When $\beta_3$ is included, its estimate is postitive, meaning that friendships between two girls with different drinking behaviors tend to form when there is third girl who is already friends with the other two. The inclusion of $\beta_3$ also leads to increases in the estimates of $\alpha_1, \alpha_2$. Thus, when modeling the transitive triplet behavior in this way, the model concludes that girls change friends more frequently. It is not clear, however, that the addition of a parameter to the objective function should effect the estimates of the rate parameters. %This strong and possibly problematic relationship between rate parameter estimates and objective function parameters is further explored in the Supporting Information. When visualizing the results of the same models fit many times to the same data, we can to see potential model behavior problems right away.   
In Figure~\ref{fig:corplots}, we further examine the relationship between each of pair of parameters common to all three models. The strongest correlation within each model is between $\beta_1$ and $\beta_2$, with absolute correlation greater than 0.90 in all three models. The $\beta_1$ parameter is also highly correlated with the $\beta_3$ parameter within model M2 ($\rho = $\Sexpr{round(corm2[5, 3])}), but it is not as highly correlated with the $\beta_4$ parameter in model M3 ($\rho = $\Sexpr{round(corM3[5, 3])}). The inherent relationship between effects is a known problem in network analysis \citep{goldenberg09}. 
% need a better reference here.

@@

The third and final model visualization principle, ``explore the process of model fitting," is crucial to understanding the CTMC models \citep{modelvis}. When fitting models in \texttt{RSiena}, several simulation processes are hidden from the user. We refer to one of these hidden pieces in the SIENA MoM algorithm as the \textit{microstep process}, which is a series of state changes in the CTMC as described in Section~\ref{sec:saoms}. At each microstep, an ego node is selected to change, and then randomly makes one change in its ties according to to the probabilities $p_{ij}$ defined in Equation~\ref{eq:pij}. This process continues until the network differs from the $x(t_m)$ by exactly $C_m$, as defined in Equation~\ref{eq:cdiff}. Between two network observations $x(t_m)$ and $x(t_{m+1})$, there can be dozens, hundreds, or even thousands of microsteps, depending on the size of the network and the value of $C_m$. We want to simulate and view these in-between steps in order to better understand the behavior of the  CTMC model. Hiding the microsteps is practical and efficient if the goal is to fit a model to data, obtain parameter estimates, and draw conclusions or make predictions. But in order to understand how the models are fit, we need to extract and explore the different steps of the process. We use these steps to visually explore the algorithms, so that we are visualizing the model fitting process, and not only the model fitting result.

@@

To visualize the underlying algorithm, we animate the microsteps that form the CTMC from $x(t_1)$ to $x(t_2)$ for the example data. The microstep animation, some frames of which are shown in Figure~\ref{fig:makinggif}, represent one possible path from $x(t_1)$ to $x(t_2)$ according to the model M1. We show each change in the network as follows: we first increase the size and change the color of the ego node to focus attention to its ties. Then, the changing tie either fades in or fades out. If there are no changes at a particular microstep, the microstep does not appear. Some frames of the animations are shown in Figure~\ref{fig:makinggif}, and the full movie, created with \texttt{tweenr} and \texttt{gganimate} can be viewed at \url{https://vimeo.com/240089108} \citep{tweenr, gganimate}. Movies similar to this animation were used to visualize the changes of dynamic networks in \citet{dynnetviz}. Frames 15-28 of Figure~\ref{fig:makinggif} show tie that is destroyed, while frames 71-84 show a tie that is created. For both changes, the ego node is emphasized with color and size. When a tie is removed, it is first emphasized with the same color and size as the node. As the animation proceeds, the tie shrinks and disappears as the ego node shrinks and changes color to match the others. If a tie is being added, it appears red from nothing, grows large for emphasis, then changes color and size to match the rest of the edges. In this network animation, we show the modeled steps of the unobserved CTMC process between two observed networks. We see each part of the hierarchical model: first, the selection of the ego node via the rate function, then the result of the actor maximizing its objective function by either deleting or adding a node. Finally, the layout of the network transitions as edges are removed or added, demonstrating how the overall network structure changes with each of the individual tie changes. Combining the tie changes in this animation makes the CTMC process the focus, demonstrating the underlying change mechanism of the dynamic network according to the model. % need more about what we learned.

@@

We continue to use animation to view the changing structure of the network via adjacency matrix visualizations. In the first adjacency matrix visualization in Figure~\ref{fig:adjmatorder}, we use the node ID to order the rows and columns. This arbitrary ordering does not easily convey the underlying structure of the network. We use \textit{matrix seriation} to reorder the rows and columns of the adjacency matrix visualization so that the structure of the network is apparent \citep{seriation}. %This lack of perceived structure would contribute to further confusion as the network chages, so we adjust the ordering so that the viewer can better perceive the structure of the network. This process is known as matrix seriation 
To seriate the matrix, we first construct a cumulative adjaceny matrix, $\mathbf{A}^{cum}$, for the series of microsteps in the CTMC simulated from $x(t_m)$ to $x(t_{m+1})$. A single entry in the cumulative adjacency matrix, $\mathbf{A}^{cum}_{ij}$, is the total number of times the edge $x_{ij}$ appears in the network from the initial observation, $x(t_m) \equiv X(0)$ to the last microstep, $X(R)$, where $R$ is the total number of microsteps taken:
\begin{equation}\label{eq:acum}
\mathbf{A}^{cum}_{ij} = \sum_{r = 0}^R X_{ij}(r). 
\end{equation}
We then perform a principal component analysis (PCA) on $\mathbf{A}^{cum}$, and use the values of the first principal component to order the vertices on the $x$ and $y$ axes for the adjacency matrix net-vis. The resulting visualization for the first network observation is shown on the right in Figure~\ref{fig:adjmatorder}. %we present the first wave of data ordered by the arbitrary node ID order alongside the seriated adjacency matrix ordered by the first principal component loading on the cumulative adjacency matrix, $\mathbf{A}^{cum}$. 
Using matrix seriation on the %PCA on $\mathbf{A}^{cum}$ to order the rows and columns of the 
adjacency matrix visualization clearly shows the two distinct connected components in the first network, which are difficult to find in the visualization ordered by node ID. We continue to use the seriated matrix in another animation of the microstep process, which is available at \url{https://vimeo.com/240092677}. Some frames of this video are shown in Figure~\ref{fig:adjmatani}. In the animation, a square appears or disappears as that edge appears or disappears in the microstep process. %Through this animation, we can see edges appearing, and then later on disappearing. 
This vanimation, unlike the node-link visualization, shows the microsteps where the chosen node does not make a change. From start to finish, we see the network evolve: sometimes ties are removed right after they are created, and sometimes several ties are removed or created in a row. The adjacency matrix animation shows the algorithm exploring the data space. In contrast to the previous animation, all possible edges are represented in the adjacency matrix visualization. By viewing this animation, we gain a better understanding of the underlying dynamics of this model. % need more oomph

<<adjmatani, fig.cap='A selection of frames from the adjacency matrix visualization animation for one series of microsteps. The node labels are removed to declutter the figure. Starting at frame 0 (the data), there are two clearly connected components: one in the bottom left, and one in the top right. By the end, the bottom left component has become more sparse, while the top right component has shrunk, but remains closely connected.', warning = FALSE, message = FALSE, results='hide', fig.height=3, out.width="\\linewidth">>=
adjmatall <- matrix(0, nrow = 16, ncol = 16)
rownames(adjmatall) <- paste0("V", 1:16)
colnames(adjmatall) <- paste0("V", 1:16)
dat2 <- sfmsall %>% filter(rep == 1)
adjmatall2 <- NULL
for(i in 0:55){
  MS <- dat2[dat2$ms == i,c("from", "to")]
  myadjmat <- adjmatall
  for(j in 1:nrow(MS)){
    myadjmat[MS[j,"from"][[1]],MS[j,"to"][[1]]] <- 1
  }
  adjmatall2 <- rbind(adjmatall2, myadjmat)
}

@@

@
\begin{figure}
\begin{subfigure}[t]{\textwidth}
<<heatmap, fig.align='center', fig.height=2.5>>=

ansnullchains %>%
  filter(period == 1) %>%
  group_by(rep) %>%
  mutate(msno = row_number()) %>%
  filter(msno == 1) %>%
  mutate(ego = as.numeric(as.character(ego))+1,
         alter = as.numeric(as.character(alter))+1,
         prob = exp(logAlterProb)) -> probsDat

ansnullchains %>%
  filter(period == 1) %>%
  group_by(rep) %>%
  mutate(msno = row_number()) %>%
  filter(msno == 1) %>%
  ungroup() %>% group_by(ego, alter) %>% 
  tally() %>% ungroup() %>% mutate(empprob = n/1000) %>% 
  mutate(ego = as.numeric(as.character(ego))+1,
         alter = as.numeric(as.character(alter))+1) ->empprobsDat

>>>>

The lack of a native coordinate system has given rise two primary net-vis methods: the node-link diagram and the adjacency matrix visualization \citep{adjmatpro}. We demonstrate these with a simpe example. Consider the network with five nodes, $\{1,2,3,4,5\}$, connected by five directed edges:  $\{2 \to 4, 3 \to 4, 1 \to 5, 3 \to 5, 5 \to 4\}$ shown in Figure~\ref{fig:toyplots}.

The first net-vis method, the node-link diagram, represents nodes with points in two dimensions and then represents edges by connecting the points with lines. These lines can also have arrows on them indicating the direction of the edge for directed networks, as shown in Figure~\ref{fig:toyplots}. Because there is often no natural placement of the nodes, they are placed in 2D at random, then adjusted with a layout algorithm.  Some commonly used algorithms, such as the Kamada-Kawai (KK) layout \citep{kamadakawai} and the Fruchterman-Reingold (FR) layout \citep{fruchterman-reingold}, are designed to mimic physical systems, drawing the graphs based on the ``forces" connecting them. For a full discussion of available node-link layout algorithms, we direct the reader to \citet{drawingalgs}. %Other algorithms use multi-dimensional scaling or properties of the adjacency matrix (e.g. its eigenstructure), to place the nodes \citep{drawingalgs}. 
In Figure~\ref{fig:toyplots}, we use the KK algorithm, and unless otherwise stated, all other node-link diagrams in this paper are laid out using the KK algorithm \citep{kamadakawai}.

@@

\subsubsection{Fitting Models to Data}\label{subsec:fitting}

@@

\noindent \textbf{Models:} To our example data, we fit three different models. Each model uses a simple rate function, $\alpha_m$, and an objective function with two or three parameters. Though our models are very simple, the model-vis techniques we present can easily be applied to much more complex models. The first model of interest, M1, contains the absolute minimum number of parameters in the objective function $f_i(x)$: 
\begin{equation}\label{eq:m1}
M1: \quad f_i(x) = \beta_1s_{i1}(x) + \beta_2s_{i2}(x),
\end{equation}
where $s_{i1}(x)$ is the density network statistic and $s_{i2}(x)$ is the reciprocity network statistic for actor $i$ at the current network state $x$ (see Figure~\ref{fig:structures}). The second and third models, M2 and M3, contain one additional parameter each in the objective function. We chose the additional parameters because they were significant ($p$-value $<$ 0.05) according to a Wald-type test (\texttt{Wald.RSiena()}) provided in the \texttt{RSiena} software \citep{RSiena}. The model M2 contains an actor-level covariate parameter, and the model M3 contains an additional strutural effect in the objective function.
\begin{align}\label{eq:m2m3}
M2: \quad f_i(x,z) & = \beta_1s_{i1}(x) + \beta_2s_{i2}(x) + \beta_3s_{i3}(x, z) \\
M3: \quad f_i(x) & = \beta_1s_{i1}(x) + \beta_2s_{i2}(x) + \beta_4s_{i4}(x),
\end{align}

@@

\noindent \textbf{Varying model settings:} The second method for obtaining collections of models is to vary model settings. We have done this already by choosing three different models M1, M2, M3 to fit to our example data sets. CTMC models, however, are much more flexible than the three simple models we have chosen. Other models could have included different rate functions for each node, where for example the rate of change could depend on drinking behavior if we have reason to believe that girls who drink more change friends more frequently. We could also model the behavior changes, not just the network changes, if we believe that friends influence behavior via peer pressure.  And as before, this method of gathering a collection of models can be used for any network model, such as the hierarchichal Bayesian network estimation model in the \texttt{R} package \texttt{sna}, which allows the user to vary hyperparameters in the prior and model parameters \citep{sna}.
% took out this next bit just because it doesn't really add much and we can just put the appendix figure into the previous section. % add more later.

@@

%\subsection{Fitting the same model to the same data}\label{sec:samemodsamedat}
\begin{figure}
\begin{subfigure}[t]{.4\textwidth}
<<distests, fig.height=7.5>>=
levels(simu2$parameter) <- c("alpha[1]","alpha[2]","beta[1]","beta[2]","beta[3]","beta[4]")
filter(simu2, parameter %in% c("beta[3]","beta[4]")) %>% 
ggplot() +
  geom_density(aes(x = estimate)) +
  facet_wrap(~parameter, scales = 'free', labeller= label_parsed, nrow = 2) +
  ThemeNoNet +
  scale_fill_brewer(palette = "Greys") + 
  xlab("estimate value") + 
  theme(legend.position = 'bottom', strip.text.x = element_text(size = 20), 
        axis.text.x = element_text(size = 20), axis.title.x = element_text(size = 20),
        axis.text.y = element_text(size = 20), axis.title.y = element_text(size = 20))
@
\caption{\label{fig:beta34} Distribution of fitted $\beta_3, \beta_4$ values.}
\end{subfigure}\hfill
\begin{subfigure}[t]{.6\textwidth}
<<img, out.width = "\\linewidth", fig.height=7.5>>=
# correlation for other params. 
simsM2 <- filter(simu2, Model == "M2") %>% spread(parameter, estimate)
corm2 <- cor(simsM2[, 3:7])
simsM3 <- filter(simu2, Model == "M3") %>% spread(parameter, estimate)
corM3 <- cor(simsM3[, 3:7])

#knitr::include_graphics("img/gray_corplot-1.png")
knitr::include_graphics("img/plot-matrix.png")

@
\caption{\label{fig:corplots} Relationships between parameters common to all models. Densities and scatterplots are colored by model. \nocite{ggally}}
\end{subfigure}
\caption{\label{fig:distests} Distribution of fitted parameter values for our three models. The inclusion of $\beta_3$ or $\beta_4$ clearly has an effect on the distributions of the rate parameters, $\alpha_1$ and $\alpha_2$, and on the other objective function parameters $\beta_1, \beta_2$.}
\end{figure}

\noindent \textbf{Fitting the model repeatedly:} The simulations required to fit the CTMC network models lead to inherent variability in the fitted values. To examine this variability, we fit models M1, M2, and M3 to the friendship network 1,000 times each. The resulting means and standard deviations of the estimates are given in Table~\ref{tab:meanests}, and the distributions of all parameter estimates are shown in Figure~\ref{fig:distests}. In Figure~\ref{fig:corplots}, we can see that the addition of $\beta_3$ to the model results in very different estimates the of the other four parameters, $\alpha_1$, $\alpha_2$, $\beta_1$, and $\beta_2$. In fact, the distributions of $\alpha_1$ and $\beta_1$ for M2 do not overlap at all with the distribution of the same parameters from model M1. When $\beta_3$ is included, its estimate is postitive, meaning that friendships between two girls with different drinking behaviors tend to form when there is third girl who is already friends with the other two. The inclusion of $\beta_3$ also leads to increases in the estimates of $\alpha_1, \alpha_2$. Thus, when modeling the transitive triplet behavior in this way, the model concludes that girls change friends more frequently. It is not clear, however, that the addition of a parameter to the objective function should effect the estimates of the rate parameters.

On the other hand, the inclusion of the indirect tie parameter, $\beta_4$, does not lead to such drastically different estimates of other model parameters.  The mean estimate of $\beta_4$ is large and negative, which suggests that indirect ties are strongly discouraged from forming. This is fairly intuitive given the context: teenage girls likely want to have direct friendships in tightly knit groups, so indirect relationships are not common.
%This strong and possibly problematic relationship between rate parameter estimates and objective function parameters is further explored in the Supporting Information. When visualizing the results of the same models fit many times to the same data, we can to see potential model behavior problems right away.   
Finally, in Figure~\ref{fig:corplots}, we demonstrate the relationship between each of pair of parameters common to all three models. The strongest correlation within each model is between $\beta_1$ and $\beta_2$, with absolute correlation greater than 0.90 in all three models. The $\beta_1$ parameter is also highly correlated with the $\beta_3$ parameter within model M2 ($\rho = $ \Sexpr{round(corm2[5, 3],3)}), but it is not as highly correlated with the $\beta_4$ parameter in model M3 ($\rho = $ \Sexpr{round(corM3[5, 3], 3)}). The inherent relationship between effects is a known problem in network analysis \citep{goldenberg09}. 
% need a better reference here.

@@

The third and final model visualization principle, ``explore the process of model fitting," is crucial to understanding the CTMC models \citep{modelvis}. When fitting models in \texttt{RSiena}, several simulation processes are hidden from the user. We refer to one of these hidden pieces in the SIENA MoM algorithm as the \textit{microstep process}, which is a series of state changes in the CTMC as described in Section~\ref{sec:saoms}. At each microstep, an ego node is selected to change, and then randomly makes one change in its ties according to to the probabilities $p_{ij}$ defined in Equation~\ref{eq:pij}. The microstep process is a key piece of the model fitting procedure and is used in the third phase mentioned in Section~\ref{subsec:fitting} \citep{SienaAlgs}. During modeling, the process proceeds until the network differs from the $x(t_m)$ by exactly $C_m$, as defined in Equation~\ref{eq:cdiff}. Between two network observations $x(t_m)$ and $x(t_{m+1})$, there can be dozens, hundreds, or even thousands of microsteps, depending on the size of the network and the value of $C_m$. We want to simulate and view these in-between steps in order to better understand the behavior of the CTMC model. The default in the software is discard the microsteps, which is practical and efficient if the goal is to fit a model to data, obtain parameter estimates, and draw conclusions or make predictions. But in order to understand how the models are fit, we need to extract and explore the different steps of the process, which is possible in \texttt{RSiena} with some small changes to the default settings. We use these steps to visually explore the algorithms, so that we are visualizing the model fitting process, and not only the model fitting result.

@@

To visualize the underlying process, we animate the microsteps that form the CTMC from $x(t_1)$ to $x(t_2)$ for the example data.  The microstep animation, some frames of which are shown in Figure~\ref{fig:makinggif}, represents one possible path from $x(t_1)$ to $x(t_2)$ according to the model M1. We show each change in the network as follows: we first increase the size and change the color of the ego node to focus attention to its ties. Then, the changing tie either fades in or fades out. If there are no changes at a particular microstep, the microstep does not appear. Some frames of the animations are shown in Figure~\ref{fig:makinggif}, and the full movie, created with \texttt{tweenr} and \texttt{gganimate} can be viewed at \url{https://vimeo.com/240089108} \citep{tweenr, gganimate}. Movies similar to this animation were used to visualize the changes of dynamic networks in \citet{dynnetviz}. Frames 15-28 of Figure~\ref{fig:makinggif} show a tie that is destroyed, while frames 71-84 show a tie that is created. For both changes, the ego node is emphasized with color and size. When a tie is removed, it is first emphasized with the same color and size as the node. As the animation proceeds, the tie shrinks and disappears as the ego node shrinks and changes color to match the others. If a tie is being added, it appears red from nothing, grows large for emphasis, then changes color and size to match the rest of the edges. In this network animation, we show the modeled steps of the unobserved CTMC process between two observed networks. We see each part of the hierarchical model: first, the selection of the ego node via the rate function, then the result of the actor maximizing its objective function by either deleting or adding a node. Finally, the layout of the network transitions as edges are removed or added, demonstrating how the overall network structure changes with each of the individual tie changes. Combining the tie changes in this animation makes the CTMC process the focus, demonstrating the underlying change mechanism of the dynamic network according to the model.

With this animation, we can see how the model behaves. Some tie changes appear to alter the overall network structure, as shown by the change in the layout of the nodes, while others seem to be inevitable, such as an added reciprocal tie between two nodes in a large connected group that do not change the layout. Other edges are created in one step and destroyed the next.  When watching the animation, we are watching the model in action, which helps us better understand its behavior. This same animation method can be applied to any network model which changes one tie at a time. For example, the Erd\"{o}s-R\'{e}nyi random graph model is a series of tie additions, and we could animate the creation of a network from this model. Unlike CTMCs, however, Erd\"{o}s-R\'{e}nyi random graph models only create ties \citep{goldenberg09}. Regardless of the network model under consideration, animating the node-link diagram to view how the network changes gives insight into the model behavior and the process of model fitting. % need more about what we learned.

@@

We continue to use animation to view the changing structure of the network via adjacency matrix visualizations. In the first adjacency matrix visualization in Figure~\ref{fig:adjmatorder}, we use the node ID to order the rows and columns. This arbitrary ordering does not easily convey the underlying structure of the network. We use \textit{matrix seriation} to reorder the rows and columns of the adjacency matrix visualization so that the structure of the network is apparent \citep{seriation}. %This lack of perceived structure would contribute to further confusion as the network chages, so we adjust the ordering so that the viewer can better perceive the structure of the network. This process is known as matrix seriation 
To seriate the matrix, we first construct a cumulative adjaceny matrix, $\mathbf{A}^{cum}$, for the series of microsteps in the CTMC simulated from $x(t_m)$ to $x(t_{m+1})$. A single entry in the cumulative adjacency matrix, $\mathbf{A}^{cum}_{ij}$, is the total number of times the edge $x_{ij}$ appears in the network from the initial observation, $x(t_m) \equiv X(0)$ to the last microstep, $X(R)$, where $R$ is the total number of microsteps taken:
\begin{equation}\label{eq:acum}
\mathbf{A}^{cum}_{ij} = \sum_{r = 0}^R X_{ij}(r). 
\end{equation}
We then perform a principal component analysis (PCA) on $\mathbf{A}^{cum}$, and use the values of the first principal component to order the vertices on the $x$ and $y$ axes for the adjacency matrix net-vis. The resulting visualization for the first network observation is shown on the right in Figure~\ref{fig:adjmatorder}. %we present the first wave of data ordered by the arbitrary node ID order alongside the seriated adjacency matrix ordered by the first principal component loading on the cumulative adjacency matrix, $\mathbf{A}^{cum}$. 
Using matrix seriation on the %PCA on $\mathbf{A}^{cum}$ to order the rows and columns of the 
adjacency matrix visualization clearly shows the two distinct connected components in the first network, which are difficult to find in the visualization ordered by node ID. We continue to use the seriated matrix in another animation of the microstep process, which is available at \url{https://vimeo.com/240092677}. Some frames of the video are shown in Figure~\ref{fig:adjmatani}. In the animation, a square appears or disappears when the edge it represents appears or disappears in the microstep process. %Through this animation, we can see edges appearing, and then later on disappearing. 
This animation, unlike the previous one, shows the microsteps where the chosen node does not make a change. Starting at frame 0 (the first network observation $x(t_1)$), there are two clearly connected components: one in the bottom left, and one in the top right. With each frame, we see the network evolve: sometimes ties are removed right after they are created, and sometimes several ties are removed or created in a row. By the end, the bottom left component has become more sparsely connected, while the top right component has shrunk, but remains closely connected. In contrast to the node-link animation, the adjacency matrix animation shows how the algorithm explore the space of all possible networks, because possible edges are represented in each frame of the adjacency matrix animation.  It is also easier to see the connected components of the graph in this animation because we have clearly grouped them with matrix seriation. As with the node-link animation, the adjacency matrix animation could also be used for other dynamic network models. Because the adjacency matrix emphasizes edges, this animation method is best for large directed networks whose structure is easier to see in an adjacency matrix visualization than in a node-link visualization \citep{adjmatviz}. Both animation methods demonstrate model behavior that would have not been seen with traditional visualizations or computations.
% need more oomph

<<adjmatani, fig.cap='A selection of frames from the adjacency matrix visualization animation for one series of microsteps. The node labels are removed to declutter the figure. Complete video available at \\url{https://vimeo.com/240092677}.', warning = FALSE, message = FALSE, results='hide', fig.height=3, out.width="\\linewidth">>=
adjmatall <- matrix(0, nrow = 16, ncol = 16)
rownames(adjmatall) <- paste0("V", 1:16)
colnames(adjmatall) <- paste0("V", 1:16)
dat2 <- sfmsall %>% filter(rep == 1)
adjmatall2 <- NULL
for(i in 0:55){
  MS <- dat2[dat2$ms == i,c("from", "to")]
  myadjmat <- adjmatall
  for(j in 1:nrow(MS)){
    myadjmat[MS[j,"from"][[1]],MS[j,"to"][[1]]] <- 1
  }
  adjmatall2 <- rbind(adjmatall2, myadjmat)
}

@@

% \begin{figure}
% \begin{subfigure}[t]{\textwidth}
% <<heatmap, fig.align='center', fig.height=2.5>>=
% 
% ansnullchains %>%
%   filter(period == 1) %>%
%   group_by(rep) %>%
%   mutate(msno = row_number()) %>%
%   filter(msno == 1) %>%
%   mutate(ego = as.numeric(as.character(ego))+1,
%          alter = as.numeric(as.character(alter))+1,
%          prob = exp(logAlterProb)) -> probsDat
% 
% ansnullchains %>%
%   filter(period == 1) %>%
%   group_by(rep) %>%
%   mutate(msno = row_number()) %>%
%   filter(msno == 1) %>%
%   ungroup() %>% group_by(ego, alter) %>% 
%   tally() %>% ungroup() %>% mutate(empprob = n/1000) %>% 
%   mutate(ego = as.numeric(as.character(ego))+1,
%          alter = as.numeric(as.character(alter))+1) ->empprobsDat
% 
% p1 <- ggplot(data = probsDat,
%        aes(y = factor(ego, levels = rev(order1)),
%              x = factor(alter, levels = order1),
%              fill = prob)) +
%   geom_tile(color = 'black') +
%   scale_fill_gradient(low = 'white', high = 'black', limits = c(0,1), name = "theoretical\nprobability", labels = c("0","",".5","","1")) +
%   labs(y = "ego", x = "alter")+
%   scale_x_discrete(position = 'top') +
%   ThemeNoNet + 
%   theme(panel.grid = element_blank(), 
%           axis.text.x.top = element_text(size=rel(.7)),
%           axis.text.y.left = element_text(size=rel(.7)),
%           legend.position = "bottom", 
%           legend.title = element_text(size=rel(.7)), 
%           legend.text = element_text(size=rel(.7)), 
%           legend.key.height = unit(.5, "cm"), 
%           legend.margin=margin(0,0,0,0),
%           legend.box.margin=margin(-10,-10,-5,-10)) +
%   coord_fixed() + 
%   ggtitle("Theoretical Probabilities") + 
%   labs(x= "alter (j)", y = "ego (i)")
% p2 <- ggplot(data = empprobsDat,
%        aes(y = factor(ego, levels = rev(order1)),
%              x = factor(alter, levels = order1),
%              fill = empprob)) +
%   geom_tile(color = 'black') +
%   scale_fill_gradient(low = 'white', high = 'black', name = "empirical\nprobability", labels = c(".02", ".04", ".06"), breaks = c(.02, .04, .06)) +
%   labs(y = "ego", x = "alter")+
%   scale_x_discrete(position = 'top') +
%   ThemeNoNet + 
%     theme(panel.grid = element_blank(), 
%           axis.text.x.top = element_text(size=rel(.7)),
%           axis.text.y.left = element_text(size=rel(.7)),
%           legend.position = "bottom", 
%           legend.title = element_text(size=rel(.7)), 
%           legend.text = element_text(size=rel(.7)), 
%           legend.key.height = unit(.5, "cm"), 
%           legend.margin=margin(0,0,0,0),
%           legend.box.margin=margin(-10,-10,-5,-10)) +
%   coord_fixed() + 
%   ggtitle("Empirical Probabilities") + 
%   labs(x= "alter (j)", y = "ego (i)")
% plot_grid(p1, p2)
% @
% \caption{\label{fig:heatmap}Adjacency matrix visualizations showing the theoretical (left) and empirical (right) transition probabilities for the first microstep taken in 1,000 simulations. There are many ties with empirical probability zero.}
% \end{subfigure}
% \begin{subfigure}[t]{\textwidth}
% <<alledgests, fig.height = 3>>=
% # get a plot of all microsteps from w1 to w2 by edge
% alledges <- expand.grid(from = paste0("V", 1:16), to = paste0("V", 1:16))
% alledges$eid <- 1:nrow(alledges)
% 
% sfmsall %>% left_join(alledges) -> sfmsall2
% 
% sfmsall2 %>% group_by(rep) %>% mutate(isWave2 = (ms == max(ms))) -> sfmsall2
% 
% sfmsall2 %>% group_by(from, to, eid) %>%
%   summarize(total =n()) %>%
%   arrange(desc(total)) %>%
%   ungroup %>%
%   mutate(plotorder = row_number()) -> sfmsallorder
% 
% sfmsall2 %>% group_by(ms, eid) %>%
%   summarise(count = n()) %>%
%   mutate(alpha = count/1000) %>%
%   arrange(ms, eid) -> sfmsall3
% 
% left_join(sfmsall3, sfmsallorder, by = "eid") -> sfmsall4
% 
% #add true wave 2 to plot
% wave2 <- na.omit(actual2)
% wave2$from <- paste0("V", wave2$from)
% wave2$to <- paste0("V", wave2$to)
% wave2 <- left_join(wave2, alledges)
% wave2 <- left_join(wave2, sfmsallorder)
% 
% sfmsall5 <- sfmsall4 %>% ungroup %>% add_row(ms = max(sfmsall4$ms) + 1,
%         eid = wave2$eid,
%         count = NA,
%         alpha = 1,
%         from = wave2$from,
%         to = wave2$to,
%         total = wave2$total,
%         plotorder = wave2$plotorder)
% sfmsall5$alphaFctr <- cut(sfmsall5$alpha, c(0,.01, .05, .10, .25, .5, .75, 1, 1.1))
% ggplot(data = sfmsall5) +
%   geom_tile(aes(x = ms, y = reorder(eid,-plotorder), fill = alphaFctr)) +
